SFJ: An implementation of Semantic Featherweight Java

### 2020-03-22 conference/paper.tex

We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:
$$
\begin{array}{l}
\synclass Polygon \ \{\ldots\}
\\
\syndecl{Triangle}{Polygon} {\ldots}
\\
\syndecl{Square}{Polygon} {\ldots}
\\
\syndecl{Rhombus}{Polygon} {\ldots}
\\
\hspace{2.5cm} \ldots
\\
\synclass Diagonal \ \{
	\\
	%\qquad Diagonal ( )\   \{\ super();\ \}
	\hspace{2.5cm} \ldots\\
	\qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
	\\
\}
\end{array}
$$
We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
The syntax of types is given by the following grammar \cite{Dardha2013,Dardha2017}:
$$
\begin{array}{llll}
\tau & ::= & \alpha \ |\ \mu
& \mbox{{\em Type term}}
\\
\alpha & ::=  & \zero \ |\ \btypes \ |\ [\wt{l:\tau}] \ |\ \alpha \ \bf{and}\ \alpha \ |\ \bf{not}\ \alpha
\qquad
& \mbox{{\em Object type} ($\alpha $-type)}
\\
\mu & ::=  & \alpha \to \alpha \ |\ \mu \ \bf{and}\ \mu \ |\ \bf{not}\ \mu
& \mbox{{\em Method type} ($\mu $-type)}
\end{array}
$$
$\alpha$-types are used to type fields and $\mu$-types are used to type methods.
Type $\zero$ is the empty type.
Type $\btypes$ denotes the \emph{basic} types, such as integers, booleans, etc.
{Record} types $[\wt{l:\tau}]$, where $\wt l$ is a sequence of disjoint labels are used to type objects.
Arrow types $\alpha \to \alpha$ are used to type methods.
The boolean types using $\bf{and}$ and $\bf{not}$ have their expected set-theoretic meanings, and $\bf{or}$ is obtained by their combination.\\
The syntax of terms is given by the following grammar and is based on the standard syntax of terms in FJ \cite{featherweight,Dardha2013,Dardha2017}:
\begin{align*}
& \mbox{\textit{Class declaration}} & L \; ::= \; & \syndecl{C}{C}{\wt{\alpha \ a};\ K; \ \wt{M}\ }\\
& \mbox{\textit{Constructor}} & K \;::=\; & C\ (\wt{\alpha\ x})\ \{\ \synsuper(\wt{x});\ \wt{\synthis.a}=\wt{x}; \}\\
& \mbox{\textit{Method declaration}} & M \; ::= \; & \alpha \ m\ (\alpha \ x)\ \{\ \synret e; \} \\
& \mbox{\textit{Expressions}} & e \; ::=\; & x\ |\  c\ |\ e.a\ |\ e.m(e) \ | \ \synnew C(\wt{e})
\end{align*}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to finite trees whose leaves are constants with no cycles.
For e.g., a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we avoid it as it is inhabitable.
%Similarly the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ would also be impossible to instantiate.
When processing the abstract syntax tree (AST) of a program in SFJ, we forbid fields to have the same type as the type of the class they are defined in. 
Secondly, we mark any classes containing fields typed with only basic types as \emph{resolved} otherwise, as \emph{unresolved}. 
This is captured by the following algorithm, which checks if the type definitions in the program are valid.
\begin{verbatim}
do
    boolean resolutionOccured = false
    for class that is unresolved:
        boolean resolved = true
        for field in class that contains a class type:
            if class type is not resolved
                resolved = false
        if resolved is true:
            set class to be resolved
            resolutionOccured = true
while resolutionOccured is true

if all classes are not resolved:
    program contains invalid type definition
\end{verbatim}
%If the types in the program are finite trees whose leaves are constants with no cycles, then at each iteration of the algorithm, either all the types are going to be resolved, or we are going to be able to resolve at least one type. 
%If we do not resolve at least one type and not all types are resolved, we know we have encountered a cycle in the type definition. 
%
Now, we can define the subtyping relation.
Building upon the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the subtyping relation by defining a map from a type to the set of its subtypes, with the property that the set of values of a subtype is included in the set of values of the type.
As a first step, we start with basic types.
\setlength{\abovedisplayskip}{5pt}
\setlength{\belowdisplayskip}{5pt}
\begin{flalign*}
    &Double  = \left\{Double, Float, Int, Short, Byte\right\} && Float = \left\{Float, Short, Byte\right\}\\
    &Long = \left\{Long, Int, Short, Byte\right\}  && Int = \left\{Int, Short, Byte\right\}\\
    &Short = \left\{Short, Byte\right\} \qquad Byte = \left\{Byte\right\} && Boolean = \left\{Boolean \right\} 
\end{flalign*}
%
Note that \emph{Int} is not a subtype of \emph{Float} as a 32-bit \emph{float} cannot represent the whole set of 32-bit \emph{integer} values accurately and therefore \emph{Int} is not fully set-contained, however this is not the case for \emph{Int} and \emph{Double}. 
Similarly, \emph{Long} is not a subtype of \emph{Double}.

@@

function checkSuperSet(class, other):
    boolean flag = true
    for field in class:
        if field contains type not in relation:
            return false
        if other does not have field:
            flag = false
        else:
            if other.field.types not fully contains field.types:
                flag = false
    for method in class:
        if method contains type not in relation:
            return false
        if other does not have method:
            flag = false
        else:
            if other.method.types not fully contains method.types:
                flag = false
    if flag == true:
        add class to other subtype relation
\end{verbatim}
Let us illustrate the subtyping algorithm with our \emph{Polygons} \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.
\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} \qquad Triangle = \left\{Triangle\right\}\\ 
    &Square = \left\{Square\right\} \qquad  Rhombus  = \left\{Rhombus\right\} \qquad
    Diagonal = \left\{Diagonal\right\} 
\end{flalign*}
Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons: 
\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\} 
\end{flalign*}
If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.

>>>>

    We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:
    $$
        \begin{array}{l}
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Square}{Polygon} {\ldots}
            \\
            \syndecl{Rhombus}{Polygon} {\ldots}
            \\
            \hspace{2.5cm} \ldots
            \\
            \synclass Diagonal \ \{
            \\
            %\qquad Diagonal ( )\   \{\ super();\ \}
            \hspace{2.5cm} \ldots \\
            \qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
            \\
            \}
        \end{array}
    $$
    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method returns the length of its longest diagonal; otherwise, if the polygon is a triangle, then there will be a type error at compile time.
\end{example}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
The syntax of types is given by the following grammar \cite{Dardha2013,Dardha2017}:

$$
    \begin{array}{llll}
        \tau   & ::=                                       & \alpha \ |\ \mu
               & \mbox{{\em Type term}}
        \\
        \alpha & ::=                                       & \zero \ |\ \btypes \ |\ [\wt{l:\tau}] \ |\ \alpha \ \bf{and}\ \alpha \ |\ \bf{not}\ \alpha
        \qquad
               & \mbox{{\em Object type} ($\alpha $-type)}
        \\
        \mu    & ::=                                       & \alpha \to \alpha \ |\ \mu \ \bf{and}\ \mu \ |\ \bf{not}\ \mu
               & \mbox{{\em Method type} ($\mu $-type)}
    \end{array}
$$

$\alpha$-types are used to type fields and $\mu$-types are used to type methods.
Type $\zero$ is the empty type.
Type $\btypes$ denotes the \emph{basic} types, such as integers, booleans, etc.
    {Record} types $[\wt{l:\tau}]$, where $\wt l$ is a sequence of disjoint labels are used to type objects.
Arrow types $\alpha \to \alpha$ are used to type methods.
The boolean types using $\bf{and}$ and $\bf{not}$ have their expected set-theoretic meanings, and $\bf{or}$ is obtained by their combination.\\
The syntax of terms is given by the following grammar and is based on the standard syntax of terms in FJ \cite{featherweight,Dardha2013,Dardha2017}:
\begin{align*}
     & \mbox{\textit{Class declaration}}  & L \; ::= \; & \syndecl{C}{C}{\wt{\alpha \ a};\ K; \ \wt{M}\ }                         \\
     & \mbox{\textit{Constructor}}        & K \;::=\;   & C\ (\wt{\alpha\ x})\ \{\ \synsuper(\wt{x});\ \wt{\synthis.a}=\wt{x}; \} \\
     & \mbox{\textit{Method declaration}} & M \; ::= \; & \alpha \ m\ (\alpha \ x)\ \{\ \synret e; \}                             \\
     & \mbox{\textit{Expressions}}        & e \; ::=\;  & x\ |\  c\ |\ e.a\ |\ e.m(e) \ | \ \synnew C(\wt{e})
\end{align*}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to finite trees whose leaves are constants with no cycles.
For e.g., a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we avoid it as it is inhabitable.
%Similarly the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ would also be impossible to instantiate.
When processing the abstract syntax tree (AST) of a program in SFJ, we forbid fields to have the same type as the type of the class they are defined in.
Secondly, we mark any classes containing fields typed with only basic types as \emph{resolved} otherwise, as \emph{unresolved}.
This is captured by the following algorithm, which checks if the type definitions in the program are valid.
\begin{verbatim}
do
    boolean resolutionOccured = false
    for class that is unresolved:
        boolean resolved = true
        for field in class that contains a class type:
            if class type is not resolved
                resolved = false
        if resolved is true:
            set class to be resolved
            resolutionOccured = true
while resolutionOccured is true

if all classes are not resolved:
    program contains invalid type definition
\end{verbatim}
%If the types in the program are finite trees whose leaves are constants with no cycles, then at each iteration of the algorithm, either all the types are going to be resolved, or we are going to be able to resolve at least one type.
%If we do not resolve at least one type and not all types are resolved, we know we have encountered a cycle in the type definition.
%
Now, we can define the subtyping relation.
Building upon the interpretation of types as sets of values to define the semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in our implementation of SFJ we keep track of the subtyping relation by defining a map from a type to the set of its subtypes, with the property that the set of values of a subtype is included in the set of values of the type.
As a first step, we start with basic types.
\setlength{\abovedisplayskip}{5pt}
\setlength{\belowdisplayskip}{5pt}
\begin{flalign*}
    &Double  = \left\{Double, Float, Int, Short, Byte\right\} && Float = \left\{Float, Short, Byte\right\}\\
    &Long = \left\{Long, Int, Short, Byte\right\}  && Int = \left\{Int, Short, Byte\right\}\\
    &Short = \left\{Short, Byte\right\} \qquad Byte = \left\{Byte\right\} && Boolean = \left\{Boolean \right\}
\end{flalign*}

Note that \emph{Int} is not a subtype of \emph{Float} as a 32-bit \emph{float} cannot represent the whole set of 32-bit \emph{integer} values accurately and therefore \emph{Int} is not fully set-contained, however this is not the case for \emph{Int} and \emph{Double}.
Similarly, \emph{Long} is not a subtype of \emph{Double}.

@@

function checkSuperSet(class, other):
    boolean flag = true
    for field in class:
        if field contains type not in relation:
            return false
        if other does not have field:
            flag = false
        else:
            if other.field.types not fully contains field.types:
                flag = false
    for method in class:
        if method contains type not in relation:
            return false
        if other does not have method:
            flag = false
        else:
            if other.method.types not fully contains method.types:
                flag = false
    if flag == true:
        add class to other subtype relation
\end{verbatim}
Let us illustrate the subtyping algorithm with our \emph{Polygons} \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.
\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} \qquad Triangle = \left\{Triangle\right\}\\
    &Square = \left\{Square\right\} \qquad  Rhombus  = \left\{Rhombus\right\} \qquad
    Diagonal = \left\{Diagonal\right\}
    >>>>>>> c2071d9f146f39da75304752181339c4518fa178
\end{flalign*}
Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons:
\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
\end{flalign*}
If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.
